---
title: "Reproducible Research Project-1"
author: "Srishti Goel"
date: "5/31/2020"
output: html_document
---
#Personal Activity Monitoring Data Analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1: Downloading and Importing Dataset


```{r}
url= "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, "rr.zip")
unzip("rr.zip") 
d = read.csv("activity.csv")
head(d)
str(d)
```

## Step 2 :  Calculating total , mean and median of steps taken each day

You can also embed plots, for example:

```{r}
library(dplyr)
d11= d %>% 
  group_by(date)%>%
  summarize(total_steps= sum(steps),
            mean= mean(steps, na.rm=TRUE), median= median(steps, na.rm=TRUE) )
head(d11)
```
## Histogram of the total number of steps taken each day

```{r, results="hide"}
library(ggplot2)
qplot( total_steps, data= d11, geom = "histogram" , 
       main ="Histogram of Total Number of steps taken at each day")
```

## Mean and median number of steps taken each day
```{r}
head(d11)
```
## Calculating average daily activity pattern

```{r}
d21= d%>%
  group_by(interval)%>%
  summarize(mean = mean(steps, na.rm= TRUE))
```

## Time series plot of the average number of steps taken
```{r}
plot(d21$interval, d21$mean, type="l")

```
## The 5-minute interval that, on average, contains the maximum number of steps
```{r}
as.numeric(filter(d21, d21$mean==max(d21$mean))[1,1])
```
## Missing Values 
```{r}
sum(is.na(d))
```
There exists missing values in our data and that may bias our decisions. For fixing this mean imputation is used to fill in the missing values for the data.

## Imputation of Missing Values through mean for that 5min interval 
Average for each 5 min interval has already been calculated in dataset d21 ignoring NA values. For filling in those, for loop is used. 

```{r}
#Creating an empty vector
nstep= c()

for(i in 1: 17568){
  if (is.na(d[i,1])==TRUE) {
    nstep[i]= filter(d21, d21$interval== d[i,3] ) [1,2]
  } else {
    nstep[i]= d$steps[i]
  }
}  
```

##  Complete Dataset with no missing data
```{r}
d31= d
d31$steps= as.numeric(nstep)
```

## Calculating total, mean ad median steps for each day 

```{r}
d32= d31%>% 
  group_by(date)%>%
  summarize (total = sum(steps),mean= mean(steps), median=median(steps) )
head(d32)
```

## Histogram of the total number of steps taken each day after missing values are imputed

```{r}
qplot(total, data=d32, geom="histogram", main="Total Number of Steps taken at each Day")
```

## Devising a factor variable to indicate whether a day is weekday or weekend 
Activity levels are expected to differ on weekend and weekdays because of office working hours.So now we will study whether this expectation is backed by emperical evidence.
As mentioned in the project hints, weekdays() is used. This function returns the which day of the week does a particular date correspond to.
 
 ```{r}
d31$date = as.Date(d31$date)
day=c()
d31$day = weekdays(d31$date)
head(d31, 4)
 ```
Now a factor variable weekend which takes value 1 if the day is a weekend and zero otherwise. 
```{r}
# Column weekend with zeroes in it
d31$weekend= rep(0) 

#Finding rows which observed on weekends
sat= grep("saturday",d31$day, ignore.case= TRUE)
sun=grep("sunday", d31$day, ignore.case=TRUE)

#Replacing weekend value with 1 using above calculated vectors sat and sun
d31[sat,5 ] = 1
d31[sun,5]=1
d31$weekend= as.factor(d31$weekend)

str(d31)
d31$steps= as.numeric(d31$steps)
```
## Plot


```{r}
d41= d31%>%
  group_by(interval, weekend)%>%
  summarize(mean= mean(steps))


levels(d41$weekend)
levels(d41$weekend)= c("Weekday", "Weekend")

qplot(interval,mean, data=d41, geom= "line",facets= weekend~., ylab="Number of steps", 
      xlab="Interval", main= "")
```

